Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC). }

### 2016-01-25 ijcar2016/body.tex
tuned for space

Second, we show that if $\conv( \varphi )$ is satisfied by a
$\extendsig{\Sigma}$-interpretation $\J$, then $\varphi$ is satisfied by a
$\Sigma$-interpretation $\I$. Since $\varphi$ is in definitional form with
respect to the functions defined by $\DDD$, it must be of the form
$\DDD \wedge \varphi_0$. 
%We argue that $\J$ satisfies a $\Sigma$-formula
%$\varphi'$ that is closed under function expansion with respect to $\DDD$, and
%where $\varphi'$ entails $\varphi_0$.
%To construct $\varphi'$, 
First, we define a sequence of $\Sigma$-literals sets $L_0 \subseteq L_1 \subseteq \cdots$
such that $\J$ satisfies $L_i \cup \absconstraints( L_i )$ for $i = 0, 1, \ldots\vthinspace$.
Since $\J$ satisfies $\conv( \varphi_0 )$,
by Lemma~\ref{lem:conv}, 
$\J$ satisfies a set of literals $L \cup \absconstraints( L )$ where $L$ is a set of $\Sigma$-literals that entail $\varphi_0$.
Let $L_0 = L$.
For each $i \geq 0$,
let $\psi_i$ be the formula $\bigwedge\, \{ \conv( \varphi_{\con{f}}[ \vec t / \vec x ] ) \mid \con{f}( \vec t ) \in \terms( L_i ),\: \con{f} \in \sfundefs{\Sigma} \}$,
where $\forallf{\con{f}} \vec x.\; \varphi_{\con{f}} \in \DDD$.
Since $\J$ satisfies $\conv( \forallf{\con{f}} \vec x.\; \varphi_{\con{f}} )$ and $\absconstraints( L_i )$,
we know that $\J$ also satisfies $\psi_i$.
Thus by Lemma~\ref{lem:conv},
$\J$ satisfies a set of literals $L \cup \absconstraints( L )$ where $L$ is a set of $\Sigma$-literals that entail $\psi_i$.
Let $L_{i+1} = L_0 \cup L$.
Let $L_\infty$ be the limit of this sequence (i.e., $\ell \in L_\infty$ if and only if
$\ell \in L_i$ for some $i$),
and let $\psi$ be the $\Sigma$-formula $\bigwedge L_\infty$.
To show that $\psi$ is closed under function expansion with respect to $\DDD$,
we first note that by construction $\psi$ entails $\psi_\infty$.
For any function symbol~$\con{f}$ and terms~$\vec t$, since $\varphi_{\con{f}}[ \vec t / \vec x ]$ does not contain function definitions,
by Corollary~\ref{cor:conv},
$\conv( \varphi_{\con{f}}[ \vec t / \vec x ] )$ entails $\varphi_{\con{f}}[ \vec t / \vec x ]$.
Thus, $\psi$ entails $\{ \varphi_{\con{f}}[ \vec t / \vec x ] \mid \con{f}( \vec t ) \in \terms( \psi ), \con{f} \in \sfundefs{\Sigma} \}$,
meaning that $\psi$ is closed under function expansion with respect to $\DDD$.
Furthermore, $\psi$ entails $\varphi_0$ since $L_0 \subseteq L_\infty$.
Since $\psi$ is a $T$-satisfiable formula that is closed under function expansion with respect to $\DDD$ and $\DDD$ is admissible, 
by definition there exists a $\Sigma$-interpretation $\I$ satisfying $\psi \wedge \DDD$, which entails $\DDD \wedge \varphi_0$,
i.e., $\varphi$.
\qed
\end{proofsketch}

The intuition of the above proof is as follows.
First, $\conv( \varphi )$ cannot be unsatisfiable when $\varphi$ is satisfiable
since any $\Sigma$-interpretation that satisfies $\varphi$ can be extended in a straightforward way to 
an $\extendsig{\Sigma}$-interpretation that satisfies $\conv( \varphi )$, by
interpreting the abstract types in the same way as the cartesian products they
abstract, thereby satisfying all existential constraints introduced by $\conv$.
Second, if a model is found for $\conv( \varphi )$,
then the constraints $\exists a : \fargtype{\con{f}}.\;\, \vecfarg{\con{f}}( a ) \teq \vec t$
occurring in $\conv( \varphi )$ ensure that this model also satisfies
a $\Sigma$-formula that is closed under function expansion and
that entails the goal of $\varphi$.
This implies the existence of a model for $\varphi$,
assuming $\DDD$ is admissible.
%For clarity, we demonstrate the consequences of this theorem by means of an example.

@@

)$, it must also satisfy\begin{rep} the formula $\psi_1$:\end{rep}

@@

a model of formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $n$ must also satisfy% the formula
~$\psi$:

>>>>

Second, we show that if $\conv( \varphi )$ is satisfied by a
$\extendsig{\Sigma}$-interpretation $\J$, then $\varphi$ is satisfied by a
$\Sigma$-interpretation $\I$. Since $\varphi$ is in definitional form with
respect to the functions defined by $\DDD$, it must be of the form
$\DDD \wedge \varphi_0$. 
%We argue that $\J$ satisfies a $\Sigma$-formula
%$\varphi'$ that is closed under function expansion with respect to $\DDD$, and
%where $\varphi'$ entails $\varphi_0$.
%To construct $\varphi'$, 
First, we define a sequence of $\Sigma$-literals sets $L_0 \subseteq L_1 \subseteq \cdots$
such that $\J$ satisfies $L_i \cup \absconstraints( L_i )$ for $i = 0, 1, \ldots\vthinspace$.
Since $\J$ satisfies $\conv( \varphi_0 )$,
by Lemma~\ref{lem:conv}, 
$\J$ satisfies a set of literals $L \cup \absconstraints( L )$ where $L$ is a set of $\Sigma$-literals that entail $\varphi_0$.
Let $L_0 = L$.
For each $i \geq 0$,
let $\psi_i$ be the formula $\bigwedge\, \{ \conv( \varphi_{\con{f}}[ \vec t / \vec x ] ) \mid \con{f}( \vec t ) \in \terms( L_i ),\: \con{f} \in \sfundefs{\Sigma} \}$,
where $\forallf{\con{f}} \vec x.\; \varphi_{\con{f}} \in \DDD$.
Since $\J$ satisfies $\conv( \forallf{\con{f}} \vec x.\; \varphi_{\con{f}} )$ and $\absconstraints( L_i )$,
we know that $\J$ also satisfies $\psi_i$.
Thus by Lemma~\ref{lem:conv},
$\J$ satisfies a set of literals $L \cup \absconstraints( L )$ where $L$ is a set of $\Sigma$-literals that entail $\psi_i$.
Let $L_{i+1} = L_0 \cup L$.
Let $L_\infty$ be the limit of this sequence (i.e., $\ell \in L_\infty$ if and only if
$\ell \in L_i$ for some $i$),
and let $\psi$ be the $\Sigma$-formula $\bigwedge L_\infty$.
To show that $\psi$ is closed under function expansion with respect to $\DDD$,
we first note that by construction $\psi$ entails $\psi_\infty$.
For any function symbol~$\con{f}$ and terms~$\vec t$, since $\varphi_{\con{f}}[ \vec t / \vec x ]$ does not contain function definitions,
by Corollary~\ref{cor:conv},
$\conv( \varphi_{\con{f}}[ \vec t / \vec x ] )$ entails $\varphi_{\con{f}}[ \vec t / \vec x ]$.
Thus, $\psi$ entails $\{ \varphi_{\con{f}}[ \vec t / \vec x ] \mid \con{f}( \vec t ) \in \terms( \psi ), \con{f} \in \sfundefs{\Sigma} \}$,
meaning that $\psi$ is closed under function expansion with respect to $\DDD$.
Furthermore, $\psi$ entails $\varphi_0$ since $L_0 \subseteq L_\infty$.
Since $\psi$ is a $T$-satisfiable formula that is closed under function expansion\begin{rep} with respect to $\DDD$\end{rep} and $\DDD$ is admissible, 
by definition there exists a $\Sigma$-interpretation $\I$ satisfying $\psi \wedge \DDD$, which entails $\DDD \wedge \varphi_0$,
i.e., $\varphi$.
\qed
\end{proofsketch}

The intuition of the above proof is as follows.
First, $\conv( \varphi )$ cannot be unsatisfiable when $\varphi$ is satisfiable
since any $\Sigma$-interpretation that satisfies $\varphi$ can be extended in a straightforward way to 
an $\extendsig{\Sigma}$-interpretation that satisfies $\conv( \varphi )$, by
interpreting the abstract types in the same way as the cartesian products they
abstract, thereby satisfying all existential constraints introduced by $\conv$.
Conversely, if a model is found for $\conv( \varphi )$,
existential constraints
%the constraints $\exists a : \fargtype{\con{f}}.\;\, \vecfarg{\con{f}}( a ) \teq \vec t$
introduced by $\conv$ ensure that this model also satisfies
a $\Sigma$-formula that is closed under function expansion and
that entails the goal of $\varphi$.
This implies the existence of a model for $\varphi$,
assuming $\DDD$ is admissible.
%For clarity, we demonstrate the consequences of this theorem by means of an example.

@@

)$, it must also satis\rlap{fy}

@@

a model of formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $n$ must also satisfy~$\psi$:
